Reasonably Programmable Syntax

### 2017-06-05 popl18/syntax-popl18.tex
finished reorganizing intro

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-12px}
\caption{An example of unreasonable program text.}
\vspace{-8px}
\label{fig:K-dialect}
\end{figure}

@@

This paper introduces a more reasonable approach: typed literal macros (TLMs). TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, whose bodies are left initially unparsed according to the fixed context free syntax of the language (i.e. they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase. TLMs come equipped with uniquely powerful abstract reasoning principles. We will more precisely characterize these abstract reasoning principles as we proceed. For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which applies a TLM to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}. Without examining the expansion of Line 3, the client programmer can reason as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have. We will see an example of a TLM definition in the next section.

Moreover, each segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

\begin{figure}
\begin{lstlisting}[numbers=none,xleftmargin=0px]
  val w = compute_w ()
  val x = compute_x ()
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-5px}
\label{fig:K-tsm-example}
\end{figure}

WE WILL CONSIDER A VARIETY OF TERM REWRITING SYSTEMS, INCLUDING PPX, EARLY MACROS, READER MACROS, TEMPLATE HASKELL, QUASIQUOTES, HYGIENIC MACROS, TYPED HYGIENIC MACROS ALA SCALA, TYPE-SPECIFIC LANGUAGES, IN SEC 6. FOR NOW, LET US JUMP IMMEDIATELY TO OUR SYSTEM OF TYPED LITERAL MACROS AND DEVELOP AN INTUITION FOR HOW THEY ADDRESS THE PROBLEMS RAISED ABOVE. THE XHTML EXAMPLE CAN BE EXPRESSED IN OUR SYSTEM AS FOLLOWS:

BUT THAT DOESNT WORK BECAUSE OF HYGIENE! THERE ARE OTHER PROBLEMS AS WELL,

For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}. We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
\begin{lstlisting}[numbers=none]
  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
\end{lstlisting}
Again, there are problems reasoning modularly and abstractly about such constructions. More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up. It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary. For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}

\subsection{Contributions}\label{sec:contributions}
This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles. % comparable to the level of control they have when defining a syntax dialect.

% Clients apply TLMs to \emph{generalized literal forms}. 
Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
\li{$html} 
 to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
 to clearly them from variables. The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TLM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}

@@

The remainder of this document is organized as follows.
\begin{itemize}
\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} and describes the process of typed expansion and expansion validation by example.
\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns that match values of a specified type. The hygiene condition for pattern TLMs is interesting.
\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.

% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala. %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
\item Sec. \ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}
\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the dictionary signature, and also over different choices of the co-domain type.
\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body. This is particularly convenient for working with abstract types \emph{a la} ML. Partial parameter application in TLM abbreviations lowers the cost of this explicit parameter-passing discipline.
\end{enumerate}
This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
\item Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
\item Sec. \ref{sec:related-work} discusses additional related work, including closely related work by \citet{TSLs,sac15} on type-specific languages. We conclude with a brief discussion of present limitations and future directions.
\end{itemize}

Due to space limitations, many of the standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. For convenience, the examples that we give are written in a full-scale functional language under development called VerseML. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

@@

\subsection{Another Existing Approach: Term Rewriting}

>>>>

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-10px}
\caption{An example of unreasonable program text.}
\vspace{-9px}
\label{fig:K-dialect}
\end{figure}

@@

This paper introduces a more reasonable approach: typed literal macros (TLMs). TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context free syntax of the language (i.e. they are initially parsed like raw string literals.) The applied TLM parses the literal body and generates an expansion during a phase known as typed expansion, which generalizes the usual typing phase. For example, Fig. \ref{fig:K-tsm-example} applies a TLM to express the ``unreasonable'' example from Fig. \ref{fig:K-dialect}.

TLMs come equipped with uniquely powerful abstract reasoning principles. We will more precisely characterize these abstract reasoning principles as we proceed. For now, let us develop some informal intuitions -- the client programmer can reason about Line 3 as follows:
\begin{enumerate}
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for typed expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax the TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can produce from it an accurate \emph{segmentation} of the literal body that distinguishes spliced segments, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. This segmentation can be communicated by a pretty printer or code editor using secondary notation, e.g. colors in this document. By examining the segmentation of Line 3, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way.
\item \textbf{(Capture)} The system avoids capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture an unseen binding of \li{x} in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope. The client is therefore free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

%\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}

\begin{figure}
\vspace{-4px}
\begin{lstlisting}[numbers=none,xleftmargin=0px]
  val w = compute_w ()
  val x = compute_x ()
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-8px}
\caption{TLMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-8px}
\label{fig:K-tsm-example}
\end{figure}

The remainder of this document is organized as follows.
\begin{itemize}
\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TLM} for XHTML and describes initial parsing, typed expansion and expansion validation in more detail.
\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TLM}, i.e. a TLM that generates patterns. %The hygiene condition for pattern TLMs is interesting.
\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.

% We next develop a more sophisticated account of TLMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TLMs into full-scale general-purpose languages like ML or Scala. %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
\item Sec. \ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}
\item They allow for TLMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, we consider a parametric TLM for dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the co-domain type.
\item Parameters also give expansions access to terms other than those explicitly spliced in. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application in TLM abbreviations lowers the cost of this parameter-passing discipline.
\end{enumerate}
This section culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with ML-style modules and parametric TLMs.
\item Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
\item Sec. \ref{sec:related-work} compares TLMs to the related work, including closely related work by \citet{TSLs,sac15} on type-specific languages. \item Sec. \ref{sec:discussion} concludes with a summary of our contributions and a brief discussion of present limitations and future research directions.
\end{itemize}

Due to space limitations, many standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. For convenience, the examples that we give are written in a full-scale functional language under development called VerseML. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

% \subsection{Contributions}\label{sec:contributions}
% This work introduces a system of \textbf{typed literal macros (TLMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles. % comparable to the level of control they have when defining a syntax dialect.

% Clients apply TLMs to \emph{generalized literal forms}.

@@

@@

Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
\li{$html} 
 to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
 to clearly them from variables. The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TLM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}

@@

\subsection{Another Existing Approach: Term Rewriting}
For example, OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some term rewriting preprocessor must rewrite \cite{ocaml-manual}. We could attempt to mark a string literal containing Ur/Web-style HTML syntax with a ppx annotation, \li{xml}, as follows:
\begin{lstlisting}[numbers=none]
  [%xml "SSTR<h1>Hello, {[first_name]}!</h1>ESTR"]
\end{lstlisting}
Again, there are problems reasoning modularly and abstractly about such constructions. More than one applied preprocessor might recognize this annotation (there are, in practice, many XML/HTML libraries), so the problems of \textbf{Responsibility} come up. It is also impossible to reason abstractly about \textbf{Segmentation}, \textbf{Capture}, \textbf{Context Dependence} and \textbf{Typing} because the code that the preprocessor generates is arbitrary. For these reasons, users of this mechanism warn that they should be used sparingly.\todo{citation}
